### 2013-03-21 doc/dissertation/document/chapters/mc_background/mc_background.tex
spn derivation dump

To demonstrate this, consider the example adapted from Mervin's work
with Wagner and others in the same area \citep{mervin_variance_2012}
and presented in Figure~\ref{fig:msod_example}.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/msod_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Overlapping domain example illustrating how domain
      overlap can reduce communication costs.}
    \textit{All particles start in the blue region of interest. The
      dashed line represents 0.5 domain overlap between domains.}}
  \label{fig:msod_example}
\end{figure}
In this example, 3 particle histories are presented emanating from the
blue region of interest. Starting with particle A, if no domain
overlap is used then the only the blue domain exists on the starting
processor. Particle A is then transported through 3 other domains
before the history ends, therefore requiring three communications to
occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
shown by the dashed line, then the starting process owns enough of the
domain such that no communications must occur in order to complete the
particle A transport process. Using 0.5 domain overlap also easily
eliminates cases such as the represented by the path of particle C. In
this case, particle C is scattering between two adjacent domains,
incurring a large latency cost for a single particle. Finally, with
particle B we observe that 0.5 domain overlap will still not eliminate
all communications. However, if 1 domain overlap were used, the entire
geometry shown in Figure~\ref{fig:msod_example} would be contained on
the source processor and therefore transport of all 3 particles
without communication would occur.

@@

Consider the example in Figure~\ref{fig:procassini_example} adapted
from Procassini's work.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/procassini_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Example illustrating how domain decomposition can
      create load balance issues in Monte Carlo.}  \textit{A domain is
      decomposed into 4 zones on 8 processors with a point source in
      the lower left zone. As the particles diffuse from the source in
      the random walk sequence as shown in the top row, their tracks
      populate the entire domain. As given in the bottom row, as the
      global percentage of particles increases in a zone, that zone's
      replication count is increased.}}
  \label{fig:procassini_example}
\end{figure}
In this example, a geometry is decomposed into 4 domains on 8
processors with a point source in the bottom left domain. To begin,
because the point source is concentrated in one domain, that domain is
replicated 5 times such the amount of work it has to do per processor
is roughly balanced with the others. As the particles begin to diffuse
away from the point source, the amount of replication is adjusted to
maintain load balance. Near the end of the simulation, the diffusion
of particles is enough that all domains have equal replication.  By
doing this, load balance is improved as each domain has approximately
equal work although each domain may represent a different spatial
location and have a differing number of histories to
compute. Compared to Wagner's work where the fission source was
distributed relatively evenly throughout the domain, fixed source
problems (and especially those that have a point-like source) like
those presented in Procassini's work will be more prone to changing
load balance requirements.

\subsubsection{Reproducible Domain Decomposed Results}
\label{subsubsec:reproducible_mc}
The 2006 work of Brunner is notable in that the Monte Carlo codes used
to implement and test the algorithms adhered to a strict policy of
generating identical results independent of domain decomposition or
domain replication as derived from the work of Gentile and colleagues
\citep{gentile_obtaining_2005}. In Gentile's work, a procedure is
given for obtaining results reproducible to machine precision for an
arbitrary number of processors and domains. Differences can arise from
using a different random number sequence in each domain and performing
a sequence of floating point operations on identical data in a
different order, leading to variations in round-off error and
ultimately a non-identical answer. They use a simple example,
recreated below in Figure~\ref{fig:gentile_example}, that illustrates
these issues.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/gentile_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Gentile's example illustrating how domain
      decomposition can create reproducibility issues in Monte Carlo.}
    \textit{Both particles A and B start in zone 1 on processor
      1. Particle A moves to zone 2 on processor 2 and scatters back
      to zone 1 while B scatters in zone 1 and remains there. A1 and
      A2 denote the track of particle A that is in zone 1 while B1 and
      B2 denote the track of particle B that is in zone 1.}}
  \label{fig:gentile_example}
\end{figure}
In this example, the domain is decomposed on two processors with each
processor owning one of the two zones. Starting with particle A, it is
born in zone 1 and is transported to zone 2 where a scattering event
occurs. Concerning the first reproducibility issue, if the same
sequence of random numbers is not used to compute the trajectory from
the new scattering event, we cannot expect to achieve the same result
if the domain were not decomposed. If the numbers are different, the
scattering event in zone 2 may keep the particle there or even eject
it from the domain if the sequence were different. The second issue is
demonstrated by adding another particle B that remains in the
domain. In this case, an efficient algorithm will transport particle A
on processor 1 until it leaves zone 1 and then transport particle
B. Particle A will not renter the domain until it has been
communicated to processor 2, processor 2 performs the transport, and
it is communicated back to processor 1. If we are doing a track-length
tally in zone 1, then we sum the tracks lengths observed in that
zone. In the single processor, single zone case particle A would be
transported in its entirety and then particle B transported. This
would result in a tally sum with the following order of operations:
$(((A1+A2)+B1)+B2)$. If we were instead to use 2 processors, we would
instead have the following order: $(((A1+B1)+B2)+A2)$. In the context
of floating point operations, we cannot expect these to have an
identical result to machine precision as round-off approximations will
differ resulting in non-commutative addition.

>>>>

To demonstrate this, consider the example adapted from Mervin's work
with Wagner and others in the same area \citep{mervin_variance_2012}
and presented in Figure~\ref{fig:msod_example}.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/mc_background/msod_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Overlapping domain example illustrating how domain
      overlap can reduce communication costs.}
    \textit{All particles start in the blue region of interest. The
      dashed line represents 0.5 domain overlap between domains.}}
  \label{fig:msod_example}
\end{figure}
In this example, 3 particle histories are presented emanating from the
blue region of interest. Starting with particle A, if no domain
overlap is used then the only the blue domain exists on the starting
processor. Particle A is then transported through 3 other domains
before the history ends, therefore requiring three communications to
occur in Brunner's algorithm. If a 0.5 domain overlap is permitted as
shown by the dashed line, then the starting process owns enough of the
domain such that no communications must occur in order to complete the
particle A transport process. Using 0.5 domain overlap also easily
eliminates cases such as the represented by the path of particle C. In
this case, particle C is scattering between two adjacent domains,
incurring a large latency cost for a single particle. Finally, with
particle B we observe that 0.5 domain overlap will still not eliminate
all communications. However, if 1 domain overlap were used, the entire
geometry shown in Figure~\ref{fig:msod_example} would be contained on
the source processor and therefore transport of all 3 particles
without communication would occur.

@@

Consider the example in Figure~\ref{fig:procassini_example} adapted
from Procassini's work.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/mc_background/procassini_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Example illustrating how domain decomposition can
      create load balance issues in Monte Carlo.}  \textit{A domain is
      decomposed into 4 zones on 8 processors with a point source in
      the lower left zone. As the particles diffuse from the source in
      the random walk sequence as shown in the top row, their tracks
      populate the entire domain. As given in the bottom row, as the
      global percentage of particles increases in a zone, that zone's
      replication count is increased.}}
  \label{fig:procassini_example}
\end{figure}
In this example, a geometry is decomposed into 4 domains on 8
processors with a point source in the bottom left domain. To begin,
because the point source is concentrated in one domain, that domain is
replicated 5 times such the amount of work it has to do per processor
is roughly balanced with the others. As the particles begin to diffuse
away from the point source, the amount of replication is adjusted to
maintain load balance. Near the end of the simulation, the diffusion
of particles is enough that all domains have equal replication.  By
doing this, load balance is improved as each domain has approximately
equal work although each domain may represent a different spatial
location and have a differing number of histories to
compute. Compared to Wagner's work where the fission source was
distributed relatively evenly throughout the domain, fixed source
problems (and especially those that have a point-like source) like
those presented in Procassini's work will be more prone to changing
load balance requirements.

\subsubsection{Reproducible Domain Decomposed Results}
\label{subsubsec:reproducible_mc}
The 2006 work of Brunner is notable in that the Monte Carlo codes used
to implement and test the algorithms adhered to a strict policy of
generating identical results independent of domain decomposition or
domain replication as derived from the work of Gentile and colleagues
\citep{gentile_obtaining_2005}. In Gentile's work, a procedure is
given for obtaining results reproducible to machine precision for an
arbitrary number of processors and domains. Differences can arise from
using a different random number sequence in each domain and performing
a sequence of floating point operations on identical data in a
different order, leading to variations in round-off error and
ultimately a non-identical answer. They use a simple example,
recreated below in Figure~\ref{fig:gentile_example}, that illustrates
these issues.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/mc_background/gentile_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Gentile's example illustrating how domain
      decomposition can create reproducibility issues in Monte Carlo.}
    \textit{Both particles A and B start in zone 1 on processor
      1. Particle A moves to zone 2 on processor 2 and scatters back
      to zone 1 while B scatters in zone 1 and remains there. A1 and
      A2 denote the track of particle A that is in zone 1 while B1 and
      B2 denote the track of particle B that is in zone 1.}}
  \label{fig:gentile_example}
\end{figure}
In this example, the domain is decomposed on two processors with each
processor owning one of the two zones. Starting with particle A, it is
born in zone 1 and is transported to zone 2 where a scattering event
occurs. Concerning the first reproducibility issue, if the same
sequence of random numbers is not used to compute the trajectory from
the new scattering event, we cannot expect to achieve the same result
if the domain were not decomposed. If the numbers are different, the
scattering event in zone 2 may keep the particle there or even eject
it from the domain if the sequence were different. The second issue is
demonstrated by adding another particle B that remains in the
domain. In this case, an efficient algorithm will transport particle A
on processor 1 until it leaves zone 1 and then transport particle
B. Particle A will not renter the domain until it has been
communicated to processor 2, processor 2 performs the transport, and
it is communicated back to processor 1. If we are doing a track-length
tally in zone 1, then we sum the tracks lengths observed in that
zone. In the single processor, single zone case particle A would be
transported in its entirety and then particle B transported. This
would result in a tally sum with the following order of operations:
$(((A1+A2)+B1)+B2)$. If we were instead to use 2 processors, we would
instead have the following order: $(((A1+B1)+B2)+A2)$. In the context
of floating point operations, we cannot expect these to have an
identical result to machine precision as round-off approximations will
differ resulting in non-commutative addition.

### 2013-03-21 doc/dissertation/document/chapters/spn_solutions/spn_solutions.tex
spn derivation dump

%        File: spn.tex
\documentclass[letterpaper,12pt]{article}
\usepackage[top=1.0in,bottom=1.0in,left=1.25in,right=1.25in]{geometry}
\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[usenames]{color}
\usepackage[
naturalnames = true, 
colorlinks = true, 
linkcolor = black,
anchorcolor = black,
citecolor = black,
menucolor = black,
urlcolor = blue
]{hyperref}
\usepackage{listings}
\usepackage{textcomp}
\definecolor{listinggray}{gray}{0.9}
\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
\lstset{
  backgroundcolor=\color{lbcolor},
  tabsize=4,
  rulecolor=,
  language=c++,
  basicstyle=\scriptsize,
  upquote=true,
  aboveskip={1.5\baselineskip},
  columns=fixed,
  showstringspaces=false,
  extendedchars=true,
  breaklines=true,
  prebreak =
  \raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
  frame=single,
  showtabs=false,
  showspaces=false,
  showstringspaces=false,
  identifierstyle=\ttfamily,
  keywordstyle=\color[rgb]{0,0,1},
  commentstyle=\color[rgb]{0.133,0.545,0.133},
  stringstyle=\color[rgb]{0.627,0.126,0.941},
}

\author{Stuart R. Slattery
  \\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
}

\date{\today}
\title{A Monte Carlo Synthetic Acceleration Method for the $SP_N$ Equations}
\begin{document}
\maketitle

%%---------------------------------------------------------------------------%%
\section{Introduction}
\label{sec:introduction}
The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, very large, or both, requiring tremendous computational
resources to generate an adequate solution. Modern deterministic
methods for large scale problems are commonly variants on the discrete
ordinates ($S_N$) method \cite{evans_denovo:_2010}. For full reactor
core neutronics simulations, the $S_N$ method requires potentially
trillions of unknown angular flux moments to be computed to achieve
good accuracy for the repsonses of interest. Other forms of the
transport problem, including the $P_N$ method, take on a simpler form
than the more common $S_N$ methods but lack in accuracy when compared
while still requiring considerable computational resources for
solutions in multiple dimensions.

@@

%%---------------------------------------------------------------------------%%
\section{Conclusion}
\label{sec:conclusion}

%%---------------------------------------------------------------------------%%
\pagebreak
\bibliographystyle{ieeetr} 
\bibliography{references}
\end{document}

>>>>

\chapter{Monte Carlo Solution Methods for the Simplified $P_N$ Equations}
\label{ch:mc_spn_solutions}

The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, very large, or both, requiring tremendous computational
resources to generate an adequate solution. Modern deterministic
methods for large scale problems are commonly variants on the discrete
ordinates ($S_N$) method \cite{evans_denovo:_2010}. For full reactor
core neutronics simulations, the $S_N$ method requires potentially
trillions of unknown angular flux moments to be computed to achieve
good accuracy for the repsonses of interest. Other forms of the
transport problem, including the $P_N$ method, take on a simpler form
than the more common $S_N$ methods but lack in accuracy when compared
while still requiring considerable computational resources for
solutions in multiple dimensions.

@@
